---
title: ""Is there a Gender wage Gap in the Miami Dade County?""
output: html_notebook
---
## Introduction
#Loasing Libraries and tyding the data
```{r}
library(readr)
library(tidyverse)
setwd("C:/Users/bhagy/Desktop")
Sal <- read_csv("Sal.csv")
race <- read_csv("race.csv")
employ1 <- read_csv("employ1.csv")
salfullrows <- read_csv("salfullrows.csv")
racefullrows <- read_csv("racefullrows.csv")
predictedrace <- read_csv("predicted race.csv")
newemploy<- drop_na(employ1)
data(newemploy)
cor.test(race$year,race$num_ppl)
```
```{r}
summary(newemploy)
print(newemploy)
```
#Timeseries 
```{r}
library(forecast)
library(readr)
library(dplyr)
newemploy$DATE <- as.Date(newemploy$DATE)
library(zoo)
tsdata <- ts(newemploy$unemployment_rate,start= 1990,end= 2018,frequency = 12)
# look at the trend
ddata <- decompose(tsdata)
plot(ddata)
plot(ddata$seasonal)
plot(ddata$trend)
class(tsdata)
plot(newemploy$unemployment_rate,type="l",main="Time vs Unemployment rate")
```
#Arima model
```{r}
mymodel <- auto.arima(tsdata)
mymodel
```
#fitting the models
```{r}
auto.arima(tsdata,ic="aic",trace=TRUE)
```
#residuals
```{r}
library(tseries)
adf.test(mymodel$residuals)
plot.ts(mymodel$residuals)
acf(ts(mymodel$residuals),main = "ACF Residual")
pacf(ts(mymodel$residuals),main = "PACF Residual")
```
#forecasting for unemployment
```{r}
myforecast <- forecast(mymodel, level =c (95), h= 5*12)
plot(myforecast,main = "Unemployment forecasting for the next 5 years",xlab = "Time", ylab ="Unemployment rate")
myforecast[["mean"]]
```
#exponential smoothing forecast
```{r}
eforecast <- ets(tsdata)
plot(forecast(eforecast,h=5*12))
```


#AVerage  for unemployment
```{r}
Unemployment_rate <- 2.845611
cat("predicted average Unemployment_rate for the next 5 years in Miami Dade County in :",Unemployment_rate)
```
##forecasting average salary male and females:   dataset:1(monthly salary)
```{r}
library(forecast)
library(readr)
newsalfullrows <- drop_na(salfullrows)
newsalfullrows$mon_wage <- as.numeric(newsalfullrows$mon_wage)
library(zoo)
newsalfullrows$year <- paste (newsalfullrows$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
ns <- ts(newsalfullrows$mon_wage,start= 1997, end= 2018,frequency = 12)
ns1 <- decompose(ns)
plot(ns1)
plot(ns1$seasonal)
plot(ns1$trend)
plot(newsalfullrows$mon_wage,type ="l",main = "monthly salary of both genders",xlab= "Year",ylab="monthly Salary")
class(ns1)
ammodel <- auto.arima(ns)
ammodel
auto.arima(ns,ic="aic",trace=TRUE)
pred_sal <- forecast(ammodel, level =c (95), h= 5*12)
plot(pred_sal,main = "Monthly Salary forecasting for the next 5years",xlab = "Time", ylab ="monthly salary")
pred_sal[["mean"]]

```
# Forecasting for male monthly salary
```{r}
library(forecast)
library(readr)
newsalfullrows <- drop_na(salfullrows)
newsalfullrows$mon_wage <- as.numeric(newsalfullrows$mon_wage)
malesalfullrows<- subset(newsalfullrows,sex == "Male",select=c(year,mon_wage))
library(zoo)
malesalfullrows$year <- paste (malesalfullrows$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
malens <- ts(malesalfullrows$mon_wage,start= 1997, end= 2018,frequency = 12)
ns2 <- decompose(malens)
plot(ns2)
plot(ns2$seasonal)
plot(ns2$trend)
plot(malesalfullrows$mon_wage,type ="l",main = "male monthly salary ",xlab= "Year",ylab="monthly Salary")
class(ns2)
ammodel2 <- auto.arima(malens)
ammodel2
auto.arima(malens,ic="aic",trace=TRUE)
pred_sal2 <- forecast(ammodel2, level =c (95), h= 5*12)
```
```{r}
plot(pred_sal2,main = "male Monthly Salary forecasting for the next 5years",xlab = "Time", ylab ="monthly salary")
pred_sal2[["mean"]]
```
#forecasting for female monthly salary
```{r}
library(forecast)
library(readr)
newsalfullrows <- drop_na(salfullrows)
newsalfullrows$mon_wage <- as.numeric(newsalfullrows$mon_wage)
femalesalfullrows<- subset(newsalfullrows,sex == "Female",select=c(year,mon_wage))
library(zoo)
femalesalfullrows$year <- paste (femalesalfullrows$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
femalens <- ts(femalesalfullrows$mon_wage,start= 1997, end= 2018,frequency = 12)
ns3 <- decompose(femalens)
plot(ns3)
plot(ns3$seasonal)
plot(ns3$trend)
plot(femalesalfullrows$mon_wage,type ="l",main = " female monthly salary",xlab= "Year",ylab="monthly Salary")
class(ns3)
ammodel3 <- auto.arima(femalens)
ammodel3
auto.arima(femalens,ic="aic",trace=TRUE)
pred_sal3 <- forecast(ammodel3, level =c (95), h= 5*12)
plot(pred_sal3,main = "female Monthly Salary forecasting for the next 5years",xlab = "Time", ylab ="monthly salary")
pred_sal3[["mean"]]
```
# calculating monthly women salary for the year 2020

```{r}
 womensal <- 2391.5799/1337.244
womensal
cat("The average mens monthly salary is ",womensal,"times of womens monthly salary")
```
#calculating how much women earns for every men who earns a dollar 
```{r}
wcents <- 1/womensal
wcents
cat("The women earns",wcents*100, "cents for every men who earns a dollar")
```
#forecasting for race 1: White Alone
```{r}
library(forecast)
library(readr)
newracefullrows <- drop_na(racefullrows)
newracefullrows$EarnS <- as.numeric(newracefullrows$EarnS)
wracefullrows<- subset(newracefullrows,race == "White Alone",select=c(year,EarnS))
wracefullrows$year <- paste (wracefullrows$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
wrc <- ts(wracefullrows$EarnS,start= 1997, end= 2018,frequency = 12)
ns4 <- decompose(wrc)
plot(ns4)
plot(ns4$seasonal)
plot(ns4$trend)
plot(wracefullrows$EarnS,type ="l",main = " whites monthly salary",xlab= "Year",ylab="monthly Salary")
class(ns4)
ammodel4 <- auto.arima(wrc)
ammodel4
auto.arima(wrc,ic="aic",trace=TRUE)
pred_sal4 <- forecast(ammodel4, level =c (95), h= 3*12)
plot(pred_sal4,main = "whites Monthly Salary forecasting for the next 3years",xlab = "Time", ylab ="monthly salary")
pred_sal4[["mean"]]
```
#forecasting for race 2: Black or African American Alone
```{r}
library(forecast)
library(readr)
newracefullrows <- drop_na(racefullrows)
newracefullrows$EarnS <- as.numeric(newracefullrows$EarnS)
bracefullrows<- subset(newracefullrows,race == "Black or African American Alone",select=c(year,EarnS))
bracefullrows$year <- paste (bracefullrows$year,"/01/01",sep ="")
brc <- ts(bracefullrows$EarnS,start= 1997, end= 2018,frequency = 12)
ns5 <- decompose(brc)
plot(ns5)
plot(ns5$seasonal)
plot(ns5$trend)
plot(bracefullrows$EarnS,type ="l",main = " Blacks monthly salary",xlab= "Year",ylab="monthly Salary")
class(ns5)
ammodel5 <- auto.arima(brc)
ammodel5
auto.arima(brc,ic="aic",trace=TRUE)
pred_sal5 <- forecast(ammodel5, level =c (95), h= 3*12)
plot(pred_sal5,main = "blacks Monthly Salary forecasting for the next 3years",xlab = "Time", ylab ="monthly salary")
pred_sal5[["mean"]]
```
#forecasting for race 3: American Indian or Alaska Native Alone
```{r}
library(forecast)
library(readr)
newracefullrows$EarnS <- as.numeric(newracefullrows$EarnS)
aracefullrows <- subset(newracefullrows,race == "American Indian or Alaska Native Alone",select=c(year,EarnS))
head(aracefullrows)
aracefullrows$year <- paste(aracefullrows$year,"/01/01",sep ="")
arc <- ts(aracefullrows$EarnS,start= 1997, end= 2018,frequency = 12)
ns6 <- decompose(arc)
plot(ns6)
plot(ns6$seasonal)
plot(ns6$trend)
plot(aracefullrows$EarnS,type ="l",main = "American Indian or Alaska Native Alone monthly salary",xlab= "Year",ylab="monthly Salary")
class(ns6)
ammodel6 <- auto.arima(arc)
ammodel6
auto.arima(arc,ic="aic",trace=TRUE)
pred_sal6 <- forecast(ammodel6, level =c (95), h= 3*12)
plot(pred_sal6,main = "American Indian Monthly Salary forecasting for the next 3years",xlab = "Time", ylab ="monthly salary")
pred_sal6[["mean"]]
```
#forecasting for race 4: Asian Alone
```{r}
library(forecast)
library(readr)
newracefullrows$EarnS <- as.numeric(newracefullrows$EarnS)
asracefullrows <- subset(newracefullrows,race == "Asian Alone",select=c(year,EarnS,race))
head(asracefullrows)
asracefullrows$year <- paste(asracefullrows$year,"/01/01",sep ="")
asrc <- ts(asracefullrows$EarnS,start= 1997, end= 2018,frequency = 12)
ns7 <- decompose(asrc)
plot(ns7)
plot(ns7$seasonal)
plot(ns7$trend)
plot(asracefullrows$EarnS,type ="l",main = "Asian Alone monthly salary",xlab= "Year",ylab="monthly Salary")
class(ns7)
ammodel7 <- auto.arima(asrc)
ammodel7
auto.arima(asrc,ic="aic",trace=TRUE)
pred_sal7 <- forecast(ammodel7, level =c (95), h= 3*12)
plot(pred_sal7,main = "Asian Alone Monthly Salary forecasting for the next 3years",xlab = "Time", ylab ="monthly salary")
pred_sal7[["mean"]]
```
#forecasting for race 5: Native Hawaiian or Other Pacific Islander Alone
```{r}
library(forecast)
library(readr)
newracefullrows$EarnS <- as.numeric(newracefullrows$EarnS)
nracefullrows <- subset(newracefullrows,race == "Native Hawaiian or Other Pacific Islander Alone",select=c(year,EarnS,race))
head(nracefullrows)
nracefullrows$year <- paste(nracefullrows$year,"/01/01",sep ="")
nrc <- ts(nracefullrows$EarnS,start= 1997, end= 2018,frequency = 12)
ns8 <- decompose(nrc)
plot(ns8)
plot(ns8$seasonal)
plot(ns8$trend)
plot(nracefullrows$EarnS,type ="l",main = "Native Hawaiian or Other Pacific Islander Alone monthly salary",xlab= "Year",ylab="monthly Salary")
class(ns8)
ammodel8 <- auto.arima(nrc)
ammodel8
auto.arima(nrc,ic="aic",trace=TRUE)
pred_sal8 <- forecast(ammodel8, level =c (95), h= 3*12)
plot(pred_sal8,main = "Native Hawaiian or Other Pacific Islander Alone Monthly Salary forecasting for the next 3years",xlab = "Time", ylab ="monthly salary")
pred_sal8[["mean"]]
```

#forecasting for race 6: Two or More Race Groups

```{r}
library(forecast)
library(readr)
newracefullrows$EarnS <- as.numeric(newracefullrows$EarnS)
tracefullrows <- subset(newracefullrows,race == "Two or More Race Groups",select=c(year,EarnS,race))
head(tracefullrows)
tracefullrows$year <- paste(tracefullrows$year,"/01/01",sep ="")
trc <- ts(tracefullrows$EarnS,start= 1997, end= 2018,frequency = 12)
ns9 <- decompose(trc)
plot(ns9)
plot(ns9$seasonal)
plot(ns9$trend)
plot(tracefullrows$EarnS,type ="l",main = "Two or More Race Groups",xlab= "Year",ylab="monthly Salary")
class(ns9)
ammodel9 <- auto.arima(trc)
ammodel9
auto.arima(trc,ic="aic",trace=TRUE)
pred_sal9 <- forecast(ammodel9, level =c (95), h= 3*12)
plot(pred_sal9,main = "Two or More Race Groups Monthly Salary forecasting for the next 3years",xlab = "Time", ylab ="monthly salary")
pred_sal9[["mean"]]
```
```{r}
ggplot(data=newsalfullrows,aes(x= factor(year),y= mon_wage,fill= sex))+geom_bar(stat="identity",position= "dodge")+labs(title="Monthly salary of both male and female",x="year",y="Monthly salary")
ggplot(data=newsalfullrows,aes(x=EmpTotal,y=mon_wage,color=sex))+geom_point()+geom_smooth(color="black")+labs(title="Monthly salary vs No of employees",x="Monthly salary", y="No of Employees")
```

```{r}
ggplot(data=newracefullrows,aes(x= year,y= EarnS,color= race))+geom_point()+geom_smooth(color="black")+scale_y_continuous(limit=c(0,20000))
ggplot(data=newracefullrows,aes(x= year,y= EarnS))+geom_bar(stat="identity",position="dodge",aes(fill= race))+scale_y_continuous(limit=c(0,20000))
```


#subsetting male average salary
```{r}
library('ggplot2')
library('forecast')
library('tseries')
#newsal <- drop_na(Sal)
newsalfullrows$year <- as.Date(newsalfullrows$year)
nsal <- subset(newsalfullrows, sex == "Male", select = c(mon_wage, year,sex,industry))
head(nsal)
dd <- aggregate(nsal$mon_wage ,by=list(dept = nsal$industry), FUN = mean)
head(dd)
sorted <- dd[order(-dd$x),]
topSorted <- sorted[c(1:3),]


#plot : Males who earn highest in top departemnts
ggplot(data =topSorted, aes(x= dept, y= x) )+geom_bar(stat = "identity",fill = "blue")+labs(x = "Department", y = "salary for Male",title = "Males who earn highest in top departemnts")
# Bottom 3 departments who earn lowest
bsorted <- dd[order(dd$x),]
botSorted <- bsorted[c(1:3),]
#plot : Males who earn lowest in bottom departemnts
ggplot(data =botSorted, aes(x= dept, y= x) )+geom_bar(stat = "identity",fill = "dark green")+labs(x = "Department", y = "salary for Male",title = "Males who earn lowest in bottom departemnts")+theme(axis.text.x = element_text(angle=45,hjust = 1))
```
#subsetting female average salary
```{r}
fnsal <- subset(newsalfullrows, sex == "Female", select = c(mon_wage, year,industry,sex))
head(fnsal)
dd1 <- aggregate(fnsal$mon_wage ,by=list(dept = fnsal$industry), FUN = mean)
head(dd1)
sorted1 <- dd1[order(-dd1$x),]
topSorted1 <- sorted1[c(1:3),]
#plot : Females who earn highest in top departemnts
ggplot(data =topSorted1, aes(x= dept, y= x) )+geom_bar(stat = "identity",fill = "pink")+ labs(x = "Department", y = "salary for Female",title = " Females who earn highest in top departemnts")
# Bottom 3 departments who earn lowest for female
bfsorted <- dd1[order(dd1$x),]
botfSorted <- bfsorted[c(1:3),]
#plot : Males who earn lowest in bottom departemnts
ggplot(data =botfSorted, aes(x= dept, y= x) )+geom_bar(stat = "identity",fill = "violet")+labs(x = "Department", y = "salary for Female",title = "females who earn lowest in bottom departemnts")+theme(axis.text.x = element_text(angle=45,hjust = 1))
```
#Subsetting for male and female who earn highest in top 8 departments
```{r}
nsal1 <- subset(newsalfullrows, sex == "Male" | sex == "Female", select = c(mon_wage, year,industry,sex))
head(nsal1)
dd2 <- aggregate(nsal1$mon_wage ,by=list(dept = nsal1$industry,sex= nsal1$sex), FUN = mean)
head(dd2)
sorted2 <- dd2[order(-dd2$x),]
topSorted2 <- sorted2[c(1:10),]
ggplot(data =topSorted2, aes(x= dept, y= x,fill= sex) )+geom_bar(stat = "identity",position ="dodge")+labs(x = "Department", y = "Average salary",title = "Males and Females who earn highest in top 8 departemnts ")+theme(axis.text.x = element_text(angle = 45,hjust=1))
sorted3 <- dd2[order(dd2$x),]
bSorted3 <- sorted3[c(1:10),]
ggplot(data =bSorted3, aes(x= dept, y= x,fill= sex) )+geom_bar(stat = "identity",position ="dodge")+labs(x = "Department", y = "Average salary",title = "Males and Females who earn lowest in top 8 departemnts ")+theme(axis.text.x = element_text(angle = 45,hjust=1))

```

```{r}
newsal <- drop_na(Sal)
```
#loading average salary male and females:dataset 2(Average salary)
```{r}
library(forecast)
library(readr)
newsal$avg_wage_ft <- as.numeric(newsal$avg_wage_ft)
library(zoo)
newsal$Date <- paste (newsal$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata1 <- ts(newsal$avg_wage_ft,frequency = 12)
tsdata1
ddata1 <- decompose(tsdata1, "multiplicative")
plot(ddata1)
plot(ddata1$seasonal)
plot(ddata1$trend)
plot(newsal$avg_wage_ft,type ="l",main = "Average salary of both genders",xlab= "Year",ylab="Average Salary")
```
```{r}
class(tsdata1)
```
#Arima model
```{r}
mymodel1 <- auto.arima(tsdata1)
mymodel1
```
#fitting model
```{r}
auto.arima(tsdata1,ic="aic",trace=TRUE)
```
#Residuals
```{r}
library(tseries)
adf.test(mymodel1$residuals)
plot.ts(mymodel1$residuals)
acf(ts(mymodel1$residuals),main = "ACF Residual")
pacf(ts(mymodel1$residuals),main = "PACF Residual")
```
#time series forecasting and average salary for both genders
```{r}
myforecast1 <- forecast(mymodel1, level =c (95), h= 3*12)
plot(myforecast1,main = "Salary forecasting for the next 3years",xlab = "Time", ylab ="Average salary")
myforecast1[["mean"]]
```
#Male average salary
```{r}
library(forecast)
library(readr)
library(zoo)
newsal <- drop_na(Sal)
msal <- subset(newsal, sex_name == "Male", select = c(avg_wage_ft, year,soc_name,sex_name))
head(msal)
msal$avg_wage_ft <- as.numeric(msal$avg_wage_ft)
msal$Date <- paste (msal$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata2 <- ts(msal$avg_wage_ft,frequency = 12)
tsdata2
ddata2 <- decompose(tsdata2)
plot(ddata2)
plot(ddata2$seasonal)
plot(ddata2$trend)
#actual plot 
plot(msal$avg_wage_ft,type ="l",main = " Male Average salary",xlab= "Year",ylab="Average Salary")
```
#class
```{r}
class(tsdata2)
```
#arima model
```{r}
mymodel2 <- auto.arima(tsdata2)
mymodel2
```
#fitting the best model
```{r}
auto.arima(tsdata2,ic="aic",trace=TRUE)
```
#Residuals
```{r}
library(tseries)
adf.test(mymodel2$residuals)
plot.ts(mymodel2$residuals)
acf(ts(mymodel2$residuals),main = "ACF Residual")
pacf(ts(mymodel2$residuals),main = "PACF Residual")
```
#forecasting male salary and mean
```{r}
myforecast2 <- forecast(mymodel2, level =c (95), h= 3*12)
plot(myforecast2,main = " Male Salary forecasting for the next 3years",xlab = "Time", ylab ="Average salary")
myforecast2[["mean"]]
```
#Female average salary 

```{r}
library(forecast)
library(readr)
library(zoo)
newsal <- drop_na(Sal)
fsal <- subset(newsal, sex_name == "Female", select = c(avg_wage_ft, year,soc_name,sex_name))
head(fsal)
fsal$avg_wage_ft <- as.numeric(fsal$avg_wage_ft)
fsal$Date <- paste (fsal$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata3 <- ts(fsal$avg_wage_ft,frequency = 12)
tsdata3
ddata3 <- decompose(tsdata3)
plot(ddata3)
plot(ddata3$seasonal)
plot(ddata3$trend)
plot(fsal$avg_wage_ft,type ="l",main = " Female Average salary",xlab= "Year",ylab="Average Salary")
```
#class

```{r}
class(tsdata3)
```
#arima model
```{r}
mymodel3 <- auto.arima(tsdata3)
mymodel3
```
#fitting model
```{r}
auto.arima(tsdata3,ic="aic",trace=TRUE)
```
#Residuals
```{r}
library(tseries)
adf.test(mymodel3$residuals)
plot.ts(mymodel3$residuals)
acf(ts(mymodel3$residuals),main = "ACF Residual")
pacf(ts(mymodel3$residuals),main = "PACF Residual")
```
#forecasting female salary and mean
```{r}
myforecast3 <- forecast(mymodel3, level =c (95), h= 3*12)
plot(myforecast3,main = " Female Salary forecasting for the next 3years",xlab = "Time", ylab ="Average salary")
myforecast3[["mean"]]
```
# calculating average women salary

```{r}
avgsal<- 53392.66/43074.50
avgsal
cat("The average mens salary is ",avgsal,"times of womens salary")
```
#calculating how much women earns for every men who earns a dollar 
```{r}
cents <- 1/avgsal
cents
cat("The women earns",cents*100, "cents for every men who earns a dollar")
```

# forecasting for all races: dataset 2
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
allrace <- subset(newrace, select = c(avg_wage, year))
head(allrace)
allrace$avg_wage <- as.numeric(allrace$avg_wage)
allrace$Date <- paste (allrace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata4 <- ts(allrace$avg_wage,frequency = 12)
tsdata4
ddata4 <- decompose(tsdata4, "multiplicative")
plot(ddata4)
plot(ddata4$seasonal)
plot(ddata4$trend)
plot(allrace$avg_wage,type ="l",main = "Average salary for all races",xlab= "Year",ylab="Average Salary")
```
#class
```{r}
class(tsdata4)
```
```{r}
mymodel4 <- auto.arima(tsdata4)
mymodel4
```
```{r}
auto.arima(tsdata4,ic="aic",trace=TRUE)
```
#residuals
```{r}
library(tseries)
adf.test(mymodel4$residuals)
plot.ts(mymodel4$residuals)
acf(ts(mymodel4$residuals),main = "ACF Residual")
pacf(ts(mymodel4$residuals),main = "PACF Residual")
```
#forecasting for all races and mean
```{r}
myforecast4 <- forecast(mymodel4, level =c (95), h= 3*12)
plot(myforecast4,main = "Salary forecasting for all races in next 3years",xlab = "Time", ylab ="Average salary")
myforecast4[["mean"]]
```


# forecasting for race 1: White
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
wrace <- subset(newrace, race_name == "White",select = c(avg_wage, year,race_name))
head(wrace)
wrace$avg_wage <- as.numeric(wrace$avg_wage)
wrace$Date <- paste (wrace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata5 <- ts(wrace$avg_wage,frequency = 12)
tsdata5
ddata5 <- decompose(tsdata5, "multiplicative")
plot(ddata5)
plot(ddata5$seasonal)
plot(ddata5$trend)
plot(wrace$avg_wage,type ="l",main = "Average salary for all whites",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata5)
```
```{r}
mymodel5 <- auto.arima(tsdata5)
mymodel5
```
```{r}
auto.arima(tsdata5,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel5$residuals)
plot.ts(mymodel5$residuals)
acf(ts(mymodel5$residuals),main = "ACF Residual")
pacf(ts(mymodel5$residuals),main = "PACF Residual")
```

```{r}
myforecast5 <- forecast(mymodel5, level =c (95), h= 3*12)
plot(myforecast5,main = "Salary forecasting for all whites in next 3years",xlab = "Time", ylab ="Average salary")
myforecast5[["mean"]]
```


# forecasting for race 2: Black or African American
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
brace <- subset(newrace, race_name == "Black or African American",select = c(avg_wage, year,race_name))
head(brace)
brace$avg_wage <- as.numeric(brace$avg_wage)
brace$Date <- paste (brace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata6 <- ts(brace$avg_wage,frequency = 12)
tsdata6
ddata6 <- decompose(tsdata6, "multiplicative")
plot(ddata6)
plot(ddata6$seasonal)
plot(ddata6$trend)
plot(brace$avg_wage,type ="l",main = "Average salary for Blacks or African Americans",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata6)
```
```{r}
mymodel6 <- auto.arima(tsdata6)
mymodel6
```
```{r}
auto.arima(tsdata6,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel6$residuals)
plot.ts(mymodel6$residuals)
acf(ts(mymodel6$residuals),main = "ACF Residual")
pacf(ts(mymodel6$residuals),main = "PACF Residual")
```

```{r}
myforecast6 <- forecast(mymodel6, level =c (95), h= 3*12)
plot(myforecast6,main = "Salary forecasting for all Blacks or African Americans in next 3years",xlab = "Time", ylab ="Average salary")
myforecast6[["mean"]]
```

# forecasting for race 3: American Indian
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
Airace <- subset(newrace, race_name == "American Indian",select = c(avg_wage, year,race_name))
head(Airace)
Airace$avg_wage <- as.numeric(Airace$avg_wage)
Airace$Date <- paste (Airace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata7 <- ts(Airace$avg_wage,frequency = 12)
tsdata7
ddata7 <- decompose(tsdata7, "multiplicative")
plot(ddata7)
plot(ddata7$seasonal)
plot(ddata7$trend)
plot(Airace$avg_wage,type ="l",main = "Average salary for Americans Indian",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata7)
```
```{r}
mymodel7 <- auto.arima(tsdata7)
mymodel7
```
```{r}
auto.arima(tsdata7,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel7$residuals)
plot.ts(mymodel7$residuals)
acf(ts(mymodel7$residuals),main = "ACF Residual")
pacf(ts(mymodel7$residuals),main = "PACF Residual")
```

```{r}
myforecast7 <- forecast(mymodel7, level =c (95), h= 3*12)
plot(myforecast7,main = "Salary forecasting for American Indian in next 3years",xlab = "Time", ylab ="Average salary")
myforecast7[["mean"]]
```

# forecasting for race 4: Alaska Native
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
Anrace <- subset(newrace, race_name == "Alaska Native",select = c(avg_wage, year,race_name))
head(Anrace)
Anrace$avg_wage <- as.numeric(Anrace$avg_wage)
Anrace$Date <- paste (Anrace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata8 <- ts(Anrace$avg_wage,frequency = 12)
tsdata8
ddata8 <- decompose(tsdata8, "multiplicative")
plot(ddata8)
plot(ddata8$seasonal)
plot(ddata8$trend)
plot(Anrace$avg_wage,type ="l",main = "Average salary for Alaska Native",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata8)
```
```{r}
mymodel8 <- auto.arima(tsdata8)
mymodel8
```
```{r}
auto.arima(tsdata8,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel8$residuals)
plot.ts(mymodel8$residuals)
acf(ts(mymodel8$residuals),main = "ACF Residual")
pacf(ts(mymodel8$residuals),main = "PACF Residual")
```

```{r}
myforecast8 <- forecast(mymodel8, level =c (95), h= 3*12)
plot(myforecast8,main = "Salary forecasting for Alaska Native in next 3years",xlab = "Time", ylab ="Average salary")
myforecast8[["mean"]]
```


# forecasting for race 5: Other Native American
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
onrace <- subset(newrace, race_name == "Other Native American",select = c(avg_wage, year,race_name))
head(onrace)
onrace$avg_wage <- as.numeric(onrace$avg_wage)
onrace$Date <- paste (onrace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata9 <- ts(onrace$avg_wage,frequency = 12)
tsdata9
ddata9 <- decompose(tsdata9, "multiplicative")
plot(ddata9)
plot(ddata9$seasonal)
plot(ddata9$trend)
plot(onrace$avg_wage,type ="l",main = "Average salary for Other Native American",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata9)
```
```{r}
mymodel9 <- auto.arima(tsdata9)
mymodel9
```
```{r}
auto.arima(tsdata9,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel9$residuals)
plot.ts(mymodel9$residuals)
acf(ts(mymodel9$residuals),main = "ACF Residual")
pacf(ts(mymodel9$residuals),main = "PACF Residual")
```

```{r}
myforecast9 <- forecast(mymodel9, level =c (95), h= 3*12)
plot(myforecast9,main = "Salary forecasting for Other Native American in next 3years",xlab = "Time", ylab ="Average salary")
myforecast9[["mean"]]
```


# forecasting for race 6: Asian
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
asrace <- subset(newrace, race_name == "Asian",select = c(avg_wage, year,race_name))
head(asrace)
asrace$avg_wage <- as.numeric(asrace$avg_wage)
asrace$Date <- paste (asrace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata10 <- ts(asrace$avg_wage,frequency = 12)
tsdata10
ddata10 <- decompose(tsdata10, "multiplicative")
plot(ddata10)
plot(ddata10$seasonal)
plot(ddata10$trend)
plot(asrace$avg_wage,type ="l",main = "Average salary for Asian",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata10)
```
```{r}
mymodel10 <- auto.arima(tsdata10)
mymodel10
```
```{r}
auto.arima(tsdata10,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel10$residuals)
plot.ts(mymodel10$residuals)
acf(ts(mymodel10$residuals),main = "ACF Residual")
pacf(ts(mymodel10$residuals),main = "PACF Residual")
```

```{r}
myforecast10 <- forecast(mymodel10, level =c (95), h= 3*12)
plot(myforecast10,main = "Salary forecasting for Asian in next 3years",xlab = "Time", ylab ="Average salary")
myforecast10[["mean"]]
```





# forecasting for race 7: Native Hawaiian or Other Pacific Islander
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
nhrace <- subset(newrace, race_name == "Native Hawaiian or Other Pacific Islander",select = c(avg_wage, year,race_name))
head(nhrace)
nhrace$avg_wage <- as.numeric(nhrace$avg_wage)
nhrace$Date <- paste (nhrace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata11 <- ts(nhrace$avg_wage,frequency = 12)
tsdata11
ddata11 <- decompose(tsdata11, "multiplicative")
plot(ddata11)
plot(ddata11$seasonal)
plot(ddata11$trend)
plot(nhrace$avg_wage,type ="l",main = "Average salary for Native Hawaiian or Other Pacific Islander",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata11)
```
```{r}
mymodel11 <- auto.arima(tsdata11)
mymodel11
```
```{r}
auto.arima(tsdata11,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel11$residuals)
plot.ts(mymodel11$residuals)
acf(ts(mymodel11$residuals),main = "ACF Residual")
pacf(ts(mymodel11$residuals),main = "PACF Residual")
```

```{r}
myforecast11 <- forecast(mymodel11, level =c (95), h= 3*12)
plot(myforecast11,main = "Salary forecasting for Native Hawaiian or Other Pacific Islander in next 3years",xlab = "Time", ylab ="Average salary")
myforecast11[["mean"]]
```



# forecasting for race 8: Two or more races
```{r}
library(forecast)
library(readr)
library(zoo)
newrace <- drop_na(race)
trace <- subset(newrace, race_name == "Two or more races",select = c(avg_wage, year,race_name))
head(trace)
trace$avg_wage <- as.numeric(trace$avg_wage)
trace$Date <- paste (trace$year,"/01/01",sep ="")
#xts(tsal$Date, as.Date(tsal$Date, format='%y%m%d'))
tsdata12 <- ts(trace$avg_wage,frequency = 12)
tsdata12
ddata12 <- decompose(tsdata12, "multiplicative")
plot(ddata12)
plot(ddata12$seasonal)
plot(ddata12$trend)
plot(trace$avg_wage,type ="l",main = "Average salary for Two or more races",xlab= "Year",ylab="Average Salary")
```

```{r}
class(tsdata12)
```
```{r}
mymodel12 <- auto.arima(tsdata12)
mymodel12
```
```{r}
auto.arima(tsdata12,ic="aic",trace=TRUE)
```

```{r}
library(tseries)
adf.test(mymodel12$residuals)
plot.ts(mymodel12$residuals)
acf(ts(mymodel12$residuals),main = "ACF Residual")
pacf(ts(mymodel12$residuals),main = "PACF Residual")
```

```{r}
myforecast12 <- forecast(mymodel12, level =c (95), h= 3*12)
plot(myforecast12,main = "Salary forecasting for Two or more races in next 3years",xlab = "Time", ylab ="Average salary")
myforecast12[["mean"]]
```

#  Bias Within Job Classes: Computer programmers

```{r}
newsal <- drop_na(Sal)
bsal <- subset(newsal, soc_name == "Computer programmers",select = c(avg_wage_ft,year,sex_name,num_ppl,soc_name))
head(bsal)
ggplot(data= bsal, mapping= aes(x= factor(year), y= avg_wage_ft,fill= sex_name))+geom_bar(stat = "identity",position = "dodge2")+labs(x= "year",y= "Average salary",title = "Year vs Average Salaryfor Computer Programmers")
ggplot(data= bsal, mapping= aes(x= factor(year),y= num_ppl,fill= sex_name))+geom_bar(stat= "identity",position = "dodge2")+labs(x= "year",y= "No. of Employees",title = "Year vs No.of Employees for Computer Programmers")
```

#Bias Within Job Classes: Mechanical engineers
```{r}
newsal <- drop_na(Sal)
mesal <- subset(newsal, soc_name == "Mechanical engineers",select = c(avg_wage_ft,year,sex_name,num_ppl,soc_name))
head(mesal)
ggplot(data= mesal, mapping= aes(x= factor(year), y= avg_wage_ft,fill= sex_name))+geom_bar(stat = "identity",position = "dodge2")+labs(x= "year",y= "Average salary",title = "Year vs Average Salary for Mechanical engineers")
ggplot(data= mesal, mapping= aes(x= factor(year),y= num_ppl,fill= sex_name))+geom_bar(stat= "identity",position = "dodge2")+labs(x= "year",y= "No. of Employees",title = "Year vs No.of Employees for Mechanical engineers")
```
#Bias Within Job Classes: Electrical & electronics engineers
```{r}
newsal <- drop_na(Sal)
eesal <- subset(newsal, soc_name == "Electrical & electronics engineers",select =
          c(avg_wage_ft,year,sex_name,num_ppl,soc_name))
head(eesal)
ggplot(data= eesal, mapping= aes(x= factor(year), y= avg_wage_ft,fill= sex_name))+geom_bar(stat = "identity",position = "dodge2")+labs(x= "year",y= "Average salary",title = "Year vs Average Salary for Electrical & electronics engineers")
ggplot(data= eesal, mapping= aes(x= factor(year),y= num_ppl,fill= sex_name))+geom_bar(stat= "identity",position = "dodge2")+labs(x= "year",y= "No. of Employees",title = "Year vs No.of Employees for Electrical & electronics engineers")
```
#Bias Within Job Classes: Civil engineers
```{r}
newsal <- drop_na(Sal)
csal <- subset(newsal, soc_name == "Civil engineers",select =
          c(avg_wage_ft,year,sex_name,num_ppl,soc_name))
head(csal)
ggplot(data= csal, mapping= aes(x= factor(year), y= avg_wage_ft,fill= sex_name))+geom_bar(stat = "identity",position = "dodge2")+labs(x= "year",y= "Average salary",title = "Year vs Average Salary for Civil engineers")
ggplot(data= csal, mapping= aes(x= factor(year),y= num_ppl,fill= sex_name))+geom_bar(stat= "identity",position = "dodge2")+labs(x= "year",y= "No. of Employees",title = "Year vs No.of Employees for Civil engineers")
```

#Bias Within Job Classes:  Aerospace engineers
```{r}
newsal <- drop_na(Sal)
asal <- subset(newsal, soc_name == "Aerospace engineers",select =
          c(avg_wage_ft,year,sex_name,num_ppl,soc_name))
head(asal)
ggplot(data= asal, mapping= aes(x= factor(year), y= avg_wage_ft,fill= sex_name))+geom_bar(stat = "identity",position = "dodge2")+labs(x= "year",y= "Average salary",title = "Year vs Average Salary for Aerospace engineers")
ggplot(data= asal, mapping= aes(x= factor(year),y= num_ppl,fill= sex_name))+geom_bar(stat= "identity",position = "dodge2")+labs(x= "year",y= "No. of Employees",title = "Year vs No.of Employees for Aerospace engineers")
```
#predicted monthly salary by races
```{r}
ggplot(data= predictedrace,aes(x= factor(Year),y=mean))+geom_bar(stat="identity",position = "dodge",aes(fill=Race))+labs(title = "Predicted mean by racial group",x="year",y="Average salary")
```
##predicted monthly salary by races
```{r}
library(plotly)
library(gganimate)
predg <- subset(predictedrace,sex== "male" | sex=="female",select=c(year1,gender_mean,sex))
p<- ggplot(data= predg,aes(x= year1,y=gender_mean))+geom_bar(stat="identity",position = "dodge",aes(fill=sex))+labs(title = "Predicted mean by gender",x="year",y="Average salary")+transition_states(sex,transition_length=2,state_length=1)+enter_fade()+exit_shrink()+ease_aes("sine-in-out")
p
```
```{r}

unem <- subset(predictedrace,select=c(year2,Unemploy))
unem<- drop_na(unem)
p<- ggplot(data= unem,aes(x= year2,y=Unemploy))+geom_bar(stat="identity", fill="Navy blue")+labs(title = "Predicted Unemployment rate",x="year",y="Unemployment rate")
p
```


